\textbf{How open are hybrid journals included in transformative agreements?}

### 2024-02-27 analysis/manuscript.Rmd
final glitches

By the end of 2023, many transformative agreements had been implemented, but the interim outcomes were mixed.
The ESAC Transformative Agreement Registry[^esacregistry], the largest source of disclosure, recorded more than 800 transformative agreements, resulting in up to 900.000 open access articles published in both fully open access and hybrid journals according to the accompanying ESAC Market Watch[^esacmarketwatch]. 
Library consortia reported increased open access volume, streamlined payment and monitoring procedures, as well as extensive utilisation of open access options by the researchers they serve [@Marques_2020; @Pinhasi_2020; @Parmhed_2023]. 
The ongoing standardisation of transformative agreements has contributed to improved transparency in terms of contracts and publisher-provided article metadata [@Marques_2019; @Pinhasi_2021].
However, with the growing trend toward transformative agreements, continued reliance on big deals is perceived as problematic, because it perpetuates market concentration [@Butler_2023; @Shu_2023]. 
Whether transformative agreements lead to reduced pricing remains uncertain [@Borrego_2023] and
a substantial transition of hybrid journals towards full open access could not be observed [@Momeni_2021; @Matthias_2019].
The focus on large commercial publishers might also increase inequality [@Klebel_2023], because transformative agreements' focus on pay to publish mainly targets institutions from high-income countries, furthering a questionable journal prestige culture [@budapest]. 
Besides, an editorial-board resignation raised concerns that transformative agreements might encourage publishers to maximize journal publication volume "without regard to quality" [@Goodin_2023].

@@

Despite these controversies around transformative agreements as a means of transitioning journal publishing to full open access, there is limited evidence available on the adoption of open access in hybrid journals, and the extent to which this can be attributed to transformative agreements.
Previous studies have focused on specific countries [@P_l_nen_2020; @Huang_2020; @Haucap_2021; @Wenaas_2022; @Taubert_2023] or publisher portfolios [@Pieper_2018; @Bakker_2024; @Fraser_2023; @Jahn_2021; @Momeni_2023], while large-scale studies relied on self-reported agreement data [@Moskovkin_2022], or used APC pricing lists [@Shu_2023]. 
In particular, data availability is a limiting factor when studying the impact of transformative agreements [@Bakker_2024], because bibliometric databases, even though many allow the retrieval of open access articles in hybrid journals, do not directly attribute them to specific transformative agreements. 
Likewise, article-level open access invoicing data are only partly available [@Jahn_2021].

@@

This study combines data from multiple publicly available sources, as shown in Figure \ref{fig:data_workflow}. 
Initially, transformative agreement data retrieved from cOAlition S Journal Checker Tool[^checkertool]  provided information about journal portfolios and participating institutions. 
After identifying hybrid journals by excluding fully open access journals, Crossref served as the primary data source for article-level metadata including Creative Commons (CC) license information to indicate open access availability on publisher websites.
To determine open access articles published through transformative agreements, first author affiliations from OpenAlex [@priem2022openalex] were subsequently linked to eligible institutions according to the transformative agreement data. 
In the following, the steps are described in detail.

@@

Comparing Crossref license coverage with OpenAlex, which re-uses open access evidence from Unpaywall, a widely used open access discovery service that also parses journal websites for open content licenses [@Piwowar_2018], highlighted ongoing challenges in identifying hybrid open access [@Mart_n_Mart_n_2018; @Jahn_2021; @Zhang_2022; @Butler_2023].
Here, `r hoaddata::cr_upw |> filter(cr_year %in% 2018:2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, cr_hybrid_total) |> pull(cr_hybrid_total) |> sum()` articles under CC license were retrieved using Crossref, while `r hoaddata::cr_upw |> filter(cr_year %in% 2018:2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, upw_hybrid_total) |> pull(upw_hybrid_total) |> sum()` articles were tagged as "hybrid" according to the OpenAlex November 2023 release, which was used throughout this study.
The largest differences concerned articles published between 2018 and 2020.
In 2022, however, Crossref and OpenAlex open access numbers differ only slightly (`r hoaddata::cr_upw |> filter(cr_year %in% 2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, cr_hybrid_total) |> pull(cr_hybrid_total) |> sum()` records using Crossref vs. `r hoaddata::cr_upw |> filter(cr_year %in% 2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, upw_hybrid_total) |> pull(upw_hybrid_total) |> sum()` in OpenAlex).
Notable differences could be observed among some publishers that presumably did not provide CC license information to Crossref, including AIP Publishing, the American Physiological Society, Emerald, and the Royal Society.
Crossref license metadata was more complete with regard to the articles published by Wiley and the American Chemical Society.
Finally, inconsistent open access status information in previous OpenAlex versions was observed [@jahn2023]. After reporting this to OpenAlex, fixing this issue was still ongoing according to the release notes, which might also explain part of the discrepancy.

@@

Throughout this mostly automated data gathering and analysis process, Tidyverse tools (Wickham et al., 2019) for the R programming language (R Core Team, 2020) were used. 
The resulting data are openly available through an R data package, hoaddata, version 0.2.91 
Following @Marwick_2018, hoaddata contains not only the datasets used in the data analysis. 
It also includes code used to compile the data by connecting it to the cloud-based Google Big Query data warehouse, where scholarly big data from Crossref, OpenAlex and Unpaywall were imported.
To increase computational reproducibility, data aggregation through hoaddata was automatically carried out using GitHub Actions, a continuous integration service.

@@

Between 2018 and 2022, `r n_hybrid_jns` out of `r n_hybrid_jns_all` hybrid journals in transformative agreements published at least one open access article under a Creative Commons license.
During this period, these hybrid journals provided open access to `r n_oa` of the `r n_total` articles, representing a five-year open access proportion of `r n_oa / n_total * 100`%.
Authors who could make use of transformative agreements at the time of publication contributed `r n_ta_oa` open access articles.

@@

Figure \ref{fig:results_overview}A shows a moderate growth in the proportion of open access articles in hybrid journals, comparing the overall open access uptake and the impact of transformative agreements on this trend.
Over the five-years period from 2018 to 2022, open access increased from `r oa_by_year |> filter(cr_year == "2018") |> pull(prop)`% (n = `r oa_by_year |> filter(cr_year == "2018") |> pull(oa)`) to `r oa_by_year |> filter(cr_year == "2022") |> pull(prop)`% (n = `r oa_by_year |> filter(cr_year == "2022") |> pull(oa)`). 
Simultaneously, the total article volume of the investigated journals increased from `r oa_by_year |> filter(cr_year == "2018") |> pull(articles)` in 2018 to `r oa_by_year |> filter(cr_year == "2022") |> pull(articles)` in 2022.

Figure \ref{fig:results_overview}B highlights that the majority of hybrid open access was made available through transformative agreements in 2021 and 2022, contributing `r oa_ta_plot_df |> filter(cr_year == "2022", type == "ta_oa") |> pull(value) / oa_by_year |> filter(cr_year == "2022") |> pull(oa) * 100`% of the total open access article volume in 2022.
However, there was also a notable increase in open access provision through other means, presumably publication fees being not invoiced through transformative agreements, which increased from `r oa_ta_plot_df |> filter(cr_year == "2018", type == "no_ta") |> mutate(prop = value / n * 100) |> pull(prop)`% (n = `r oa_ta_plot_df |> filter(cr_year == "2018", type == "no_ta") |> pull(value)`) in 2018 to  `r oa_ta_plot_df |> filter(cr_year == "2022", type == "no_ta") |> mutate(prop = value / n * 100) |> pull(prop)`% (n = `r oa_ta_plot_df |> filter(cr_year == "2022", type == "no_ta") |> pull(value)`). 
<!-- This suggests that publishers were able to gain equally from individual and institutional open access publishing options.-->

@@

```{r}
jn_ind_jct <- hoaddata::jct_hybrid_jns |>
  inner_join(hoaddata::jn_ind, by = "issn_l") |>
  filter(issn_l != "0027-8424")

ta_active <- ta_oa_inst |>
  filter(issn_l != "0027-8424") |>
  filter(ta_active == TRUE, !is.na(cc)) |>
  group_by(cr_year, issn_l, esac_publisher) |>
  summarise(ta_oa = n_distinct(doi)) |>
  mutate(cr_year = as.character(cr_year))

@@

The varying degrees of adoption of open access across the three major publishers can be attributed to distinct approaches to transformative agreements. 
Springer Nature, for example, started offering open access agreements for its hybrid journal portfolio to selected consortia such as the Max Planck Society, the Swedish Bibsam Consortium and the Finnish FinELib consortium in 2015 under the name Springer Compact[^springercompact]. 
Prior to this, Springer discontinued its hybrid OA pilot agreements with a small number of institutions, including the University of Göttingen [@Schmidt_2012].
However, the Springer Compact agreements were not included in the data as they concluded prior to the start of the transformative agreement data collection in June 2021. 
Nonetheless, the results demonstrate the importance of agreements for Springer Nature's hybrid open access business over the past five years (Figure 2B).
In 2022, `r publisher_summary_df |> filter(cr_year == "2022", publisher == "Springer Nature", type == "ta_oa") |> pull(value) / publisher_summary_df |> filter(cr_year == "2022", publisher == "Springer Nature") |> pull(value) |> sum()*100`% (n = `r  publisher_summary_df |> filter(cr_year == "2022", publisher == "Springer Nature", type == "ta_oa") |> pull(value)`) of open access in Springer Nature hybrid journals was enabled through transformative agreements.
In the same year, `r publisher_summary_df |> filter(cr_year == "2022", publisher == "Wiley", type == "ta_oa") |> pull(value) / publisher_summary_df |> filter(cr_year == "2022", publisher == "Wiley") |> pull(value) |> sum()*100`% (n = `r publisher_summary_df |> filter(cr_year == "2022", publisher == "Wiley", type == "ta_oa") |> pull(value)`) of Wiley's open access articles could be linked to transformative agreements in 2022. 
By contrast, Elsevier published fewer than half of its open access articles through transformative agreements (n = `r publisher_summary_df |> filter(cr_year == "2022", publisher == "Elsevier", type == "ta_oa") |> pull(value) `; `r publisher_summary_df |> filter(cr_year == "2022", publisher == "Elsevier", type == "ta_oa") |> pull(value) / publisher_summary_df |> filter(cr_year == "2022", publisher == "Elsevier") |> pull(value) |> sum()*100`%).

@@

Table \ref{tab:subject_summary_table} presents a high-level overview of hybrid open access by AJCS subject area using fractionalised counting to account for journals belonging to more than one category. 
Between 2018 and 2022, most hybrid journals with at least one open article could be attributed to the Social Sciences category, which also includes Arts and Humanities.
However, these journals published the fewest number of articles, whereas Physical Sciences journals recorded the most articles, followed by Health Sciences and Life Sciences. 
In terms of open access,  Physical Sciences journals accounted for more than one-third of the articles published in the five-year period, followed by Health Sciences, Social Sciences and Life Sciences.

@@

Figure \ref{fig:subject_panel} presents the relative growth of hybrid open access by subject area between 2018-2022. In particular, Social Sciences inbcluding Arts and Humanties journals accounted for the strongest growth in the five-years period from `r subject_trend_plot$data |> filter(cr_year == "2018", top_level == "Social Sciences") |> pull(prop) |> sum() * 100`% (n = `r subject_trend_plot$data |> filter(cr_year == "2018", top_level == "Social Sciences") |> pull(value) |> sum()`) to `r subject_trend_plot$data |> filter(cr_year == "2022", top_level == "Social Sciences") |> pull(prop) |> sum() *100`% (n = `r subject_trend_plot$data |> filter(cr_year == "2022", top_level == "Social Sciences") |> pull(value) |> sum()`), followed by the Life Sciences from `r subject_trend_plot$data |> filter(cr_year == "2018", top_level == "Life Sciences") |> pull(prop) |> sum() * 100`% (n = `r subject_trend_plot$data |> filter(cr_year == "2018", top_level == "Life Sciences") |> pull(value) |> sum()`) to `r subject_trend_plot$data |> filter(cr_year == "2022", top_level == "Life Sciences") |> pull(prop) |> sum() *100`% (n = `r subject_trend_plot$data |> filter(cr_year == "2022", top_level == "Life Sciences") |> pull(value) |> sum()`) , Health Sciences from `r subject_trend_plot$data |> filter(cr_year == "2018", top_level == "Health Sciences") |> pull(prop) |> sum() * 100`% (n = `r subject_trend_plot$data |> filter(cr_year == "2018", top_level == "Health Sciences") |> pull(value) |> sum()`) to `r subject_trend_plot$data |> filter(cr_year == "2022", top_level == "Health Sciences") |> pull(prop) |> sum() *100`% (n = `r subject_trend_plot$data |> filter(cr_year == "2022", top_level == "Health Sciences") |> pull(value) |> sum()`) and Physical Sciences from `r subject_trend_plot$data |> filter(cr_year == "2018", top_level == "Physical Sciences") |> pull(prop) |> sum() * 100`% (n = `r subject_trend_plot$data |> filter(cr_year == "2018", top_level == "Physical Sciences") |> pull(value) |> sum()`) to `r subject_trend_plot$data |> filter(cr_year == "2022", top_level == "Physical Sciences") |> pull(prop) |> sum() *100`% (n = `r subject_trend_plot$data |> filter(cr_year == "2022", top_level == "Physical Sciences") |> pull(value) |> sum()`). 
<!-- In particular, journals in the subject areas recorded large high rate (see supplement)-->
Growth in Social Sciences category can be largely attributed to transformative agreements. 
In 2022, two-thirds of open access articles (`r subject_ta_trend_plot$data |> filter(cr_year == "2022", top_level == "Social Sciences") |> pull(prop) * 100`%, n = `r subject_ta_trend_plot$data |> filter(cr_year == "2022", top_level == "Social Sciences") |> pull(ta_oa_frac)`) were published by the first authors affiliated with participating institutions (see \ref{fig:subject_panel}B).

@@

Figure \ref{fig:subject_panel}C shows that this trend was consistent across hybrid journals belonging to the ASJC Social Sciences category.
In 2022, 25% of Social Sciences journals provided open access to at least every fourth article exclusively through transformative agreements.
However, hybrid open access through transformative agreements played a comparably lesser role in Life Sciences and Health Sciences. 
In these two subject areas, only about half of the open access articles can be linked to these agreements, both overall and on a median average across journals.
In contrast, the majority of Physical Science Journals show an increase of open access through transformative agreements compared to other options to publish open access in hybrid journals.

```{r coutnry_data_prep}
country_all <- hoaddata::jn_aff |>
  filter(!cr_year %in% c(2017, 2023), issn_l != "0027-8424") |>
  # only with cc
  filter(issn_l %in% hoaddata::cc_articles$issn_l) |>
  distinct(issn_l, cr_year, country_code, articles_total) |>
  group_by(cr_year, country_code) |>
  summarise(all = sum(articles_total))

country_oa <- hoaddata::cc_articles |>
  filter(!cr_year %in% c(2017, 2023), issn_l != "0027-8424") |>
  group_by(cr_year, country_code) |>
  summarise(oa = n_distinct(doi))

country_codes <- readr::read_csv(here::here("data", "ror_country_codes.csv"))
country_ta <-  ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  left_join(country_codes) |>
  group_by(cr_year, country_code) |>
  summarise(ta_oa = n_distinct(doi))

@@

```{r, country_plot, fig.asp=0.8}
country_plot <- ggplot(top_20_plot_df, aes(
    gsub("^20", "'", as.character(cr_year)),
    value / all_total,
    group = type,
    fill = type
  )) +
  geom_area(color = "#000000",
            linewidth = 0.15,
            stat = "identity") +
scale_fill_manual(
    NULL,
    values = c(no_ta = "#b3b3b3a0", ta_oa = "#56B4E9"),
    labels = c(no_ta = "Other", ta_oa = "Agreement"),
    guide = guide_legend(reverse = TRUE)
  ) +
  facet_wrap( ~ country_name_fct,
              nrow = 3,
              labeller = as_labeller(country_facet_labels)) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1L),
    expand = expansion(mult = c(0, 0.05)),
    breaks = scales::breaks_extended(4)
  ) +
  theme_minimal() +
  labs(y = "Hybrid OA Uptake", x = "2018 - 2022") +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.direction = "horizontal",
    axis.text.x = element_blank(),
    panel.grid.minor = element_blank(),
                panel.border = element_rect(color = "grey50", fill = NA),
    axis.title=element_text(size = 10),
    strip.text = ggtext::element_markdown(size = 7),
    plot.subtitle = ggtext::element_markdown(hjust = 0.5),
    panel.spacing = unit(1, "lines")
  )
```

@@

Between 2018 and 2022, high-income countries almost exclusively dominated hybrid open access publishing through transformative agreements. 
During this period, first-authors affiliated with institutions from Organisation for Economic Co-operation and Development (OECD) member countries published `r oecd_stats_global |> filter(cat == "OECD") |> pull(oa)` open access articles in hybrid journals, representing `r oecd_stats_global |> filter(cat == "OECD") |> pull(oa) / n_oa * 100`% of the investigated open access articles.
This disparity between OECD nations and other countries becomes even more evident when considering open access through transformative agreements, as  `r oecd_stats_global |> filter(cat == "OECD") |> pull(ta_oa)` of `r n_ta_oa`, or `r oecd_stats_global |> filter(cat == "OECD") |> pull(ta_oa) / n_ta_oa*100`% of open access articles were associated with such agreements.

>>>>

By the end of 2023, many transformative agreements had been implemented, but the interim outcomes were mixed.
The ESAC Transformative Agreement Registry[^esacregistry], the largest source of disclosure, recorded more than 800 transformative agreements, resulting in up to 900.000 open access articles published in both fully open access and hybrid journals according to the accompanying ESAC Market Watch[^esacmarketwatch]. 
Library consortia reported increased open access volume, streamlined payment and monitoring procedures, as well as extensive utilisation of open access options by the researchers they serve [@Marques_2020; @Pinhasi_2020; @Parmhed_2023]. 
The ongoing standardisation of transformative agreements has contributed to improved transparency in terms of contracts and publisher-provided article metadata [@Marques_2019; @Pinhasi_2021].
However, with the growing trend toward transformative agreements, continued reliance on big deals is perceived as problematic, because it perpetuates market concentration [@Butler_2023; @Shu_2023]. 
Whether transformative agreements lead to reduced pricing remains uncertain [@Borrego_2023] and
a substantial transition of hybrid journals towards full open access could not be observed [@Momeni_2021; @Matthias_2019].
The focus on large commercial publishers might also increase inequality [@Ross_Hellauer_2022], because transformative agreements' focus on pay to publish mainly targets institutions from high-income countries, furthering a questionable journal prestige culture [@budapest]. 
Besides, an editorial-board resignation raised concerns that transformative agreements might encourage publishers to maximize journal publication volume "without regard to quality" [@Goodin_2023].

@@

Despite these controversies around transformative agreements as a means of transitioning journal publishing to full open access, there is limited evidence available on the adoption of open access in hybrid journals, and the extent to which this can be attributed to transformative agreements.
Previous studies have focused on specific countries [@P_l_nen_2020; @Huang_2020; @Haucap_2021; @Wenaas_2022; @Taubert_2023] or publisher portfolios [@Pieper_2018; @Bakker_2024; @Fraser_2023; @Jahn_2021; @Momeni_2023; @Schmal_2024], while large-scale studies relied on self-reported agreement data [@Moskovkin_2022], or used APC pricing lists [@Shu_2023]. 
In particular, data availability is a limiting factor when studying the impact of transformative agreements [@Bakker_2024], because bibliometric databases, even though many allow the retrieval of open access articles in hybrid journals, do not directly attribute them to specific transformative agreements. 
Likewise, article-level open access invoicing data are only partly available [@Jahn_2021].

@@

This study combines data from multiple publicly available sources, as shown in Figure \ref{fig:data_workflow}. 
Initially, transformative agreement data retrieved from cOAlition S Journal Checker Tool[^checkertool]  provided information about journal portfolios and participating institutions. 
After identifying hybrid journals by excluding fully open access journals, Crossref [@Hendricks_2020] served as the primary data source for article-level metadata including Creative Commons (CC) license information to indicate open access availability on publisher websites.
To determine open access articles published through transformative agreements, first author affiliations from OpenAlex [@priem2022openalex] were subsequently linked to eligible institutions according to the transformative agreement data. 
In the following, the steps are described in detail.

@@

Comparing Crossref license coverage with OpenAlex, which re-uses open access evidence from Unpaywall, a widely used open access discovery service that also parses journal websites for open content licenses [@Piwowar_2018], highlighted ongoing challenges in identifying hybrid open access [@Mart_n_Mart_n_2018; @Jahn_2021; @Zhang_2022; @Butler_2023].
For the purpose of this study, `r hoaddata::cr_upw |> filter(cr_year %in% 2018:2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, cr_hybrid_total) |> pull(cr_hybrid_total) |> sum()` articles under CC license were retrieved using Crossref, while `r hoaddata::cr_upw |> filter(cr_year %in% 2018:2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, upw_hybrid_total) |> pull(upw_hybrid_total) |> sum()` articles were tagged as "hybrid" according to the OpenAlex November 2023 release, which was used throughout this study.
The largest differences concerned articles published between 2018 and 2020.
In 2022, however, Crossref and OpenAlex open access numbers differ only slightly (`r hoaddata::cr_upw |> filter(cr_year %in% 2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, cr_hybrid_total) |> pull(cr_hybrid_total) |> sum()` records using Crossref vs. `r hoaddata::cr_upw |> filter(cr_year %in% 2022, cat == "Global", issn_l != "0027-8424") |> distinct(issn_l, cr_year, upw_hybrid_total) |> pull(upw_hybrid_total) |> sum()` in OpenAlex).
Notable differences could be observed among some publishers that presumably did not provide CC license information to Crossref, including AIP Publishing, the American Physiological Society, Emerald, and the Royal Society.
Crossref license metadata was more complete with regard to the articles published by Wiley and the American Chemical Society.
Finally, inconsistent open access status information in previous OpenAlex versions was observed [@jahn2023]. After reporting this to OpenAlex, fixing this issue was still ongoing according to the release notes, which might also explain part of the discrepancy.

@@

Throughout this mostly automated data gathering and analysis process, Tidyverse tools (Wickham et al., 2019) for the R programming language (R Core Team, 2020) were used. 
The resulting data are openly available through an R data package, hoaddata [@hoaddata]. 
Following @Marwick_2018, hoaddata contains not only the datasets used in the data analysis. 
It also includes code used to compile the data by connecting it to the cloud-based Google Big Query data warehouse, where scholarly big data from Crossref, OpenAlex and Unpaywall were imported, using bigrquery [@bigrquery].
To increase computational reproducibility, data aggregation through hoaddata was automatically carried out using GitHub Actions, a continuous integration service.

@@

Between 2018 and 2022, `r n_hybrid_jns` out of `r n_hybrid_jns_all` hybrid journals in transformative agreements published at least one open access article under a Creative Commons license.
During this period, these hybrid journals provided open access to `r n_oa` of the `r n_total` articles, representing a five-year open access proportion of `r n_oa / n_total * 100`%.
Authors who could make use of transformative agreements at the time of publication contributed `r n_ta_oa` open access articles.
Overall, `r n_tas_with_oa` transformative agreements recorded at least one open access article from an eligible institution.

@@

Figure \ref{fig:results_overview}A shows a moderate growth in the proportion of open access articles in hybrid journals, comparing the overall open access uptake and the impact of transformative agreements on this trend.
Over the five-years period from 2018 to 2022, open access increased from `r oa_by_year |> filter(cr_year == "2018") |> pull(prop)`% (n = `r oa_by_year |> filter(cr_year == "2018") |> pull(oa)`) to `r oa_by_year |> filter(cr_year == "2022") |> pull(prop)`% (n = `r oa_by_year |> filter(cr_year == "2022") |> pull(oa)`). 
Simultaneously, the total article volume of the investigated journals rose from `r oa_by_year |> filter(cr_year == "2018") |> pull(articles)` in 2018 to `r oa_by_year |> filter(cr_year == "2022") |> pull(articles)` in 2022.

Figure \ref{fig:results_overview}B highlights that the majority of hybrid open access was made available through transformative agreements in 2021 and 2022, contributing `r oa_ta_plot_df |> filter(cr_year == "2022", type == "ta_oa") |> pull(value) / oa_by_year |> filter(cr_year == "2022") |> pull(oa) * 100`% of the total open access article volume in 2022.
However, there was also a notable increase in open access provision through other means, presumably publication fees being not invoiced through transformative agreements, from `r oa_ta_plot_df |> filter(cr_year == "2018", type == "no_ta") |> mutate(prop = value / n * 100) |> pull(prop)`% (n = `r oa_ta_plot_df |> filter(cr_year == "2018", type == "no_ta") |> pull(value)`) in 2018 to  `r oa_ta_plot_df |> filter(cr_year == "2022", type == "no_ta") |> mutate(prop = value / n * 100) |> pull(prop)`% (n = `r oa_ta_plot_df |> filter(cr_year == "2022", type == "no_ta") |> pull(value)`). 
<!-- This suggests that publishers were able to gain equally from individual and institutional open access publishing options.-->

@@

```{r}
jn_ind_jct <- hoaddata::jct_hybrid_jns |>
  inner_join(hoaddata::jn_ind, by = "issn_l") |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  filter(issn_l %in% hoaddata::cc_articles$issn_l)

ta_active <- ta_oa_inst |>
  filter(issn_l %in% jn_ind_jct$issn_l) |>
  filter(ta_active == TRUE, !is.na(cc)) |>
  group_by(cr_year, issn_l, esac_publisher) |>
  summarise(ta_oa = n_distinct(doi)) |>
  mutate(cr_year = as.character(cr_year))

@@

The varying degrees of adoption of open access across the three major publishers can be attributed to distinct approaches to transformative agreements. 
Springer Nature, for example, started offering open access agreements for its hybrid journal portfolio to selected consortia such as the Max Planck Society, the Swedish Bibsam Consortium and the Finnish FinELib consortium in 2015 under the name Springer Compact[^springercompact]. 
Prior to this, Springer had some pilot agreements with a small number of institutions, including the University of Göttingen [@Schmidt_2012].
However, the Springer Compact agreements were not included in the data as they concluded before the start of the transformative agreement data collection in June 2021. 
Nonetheless, the results demonstrate the importance of agreements for Springer Nature's hybrid open access business over the past five years (Figure 2B).
In 2022, `r publisher_summary_df |> filter(cr_year == "2022", publisher == "Springer Nature", type == "ta_oa") |> pull(value) / publisher_summary_df |> filter(cr_year == "2022", publisher == "Springer Nature") |> pull(value) |> sum()*100`% (n = `r  publisher_summary_df |> filter(cr_year == "2022", publisher == "Springer Nature", type == "ta_oa") |> pull(value)`) of open access in Springer Nature hybrid journals was enabled through transformative agreements.
In the same year, `r publisher_summary_df |> filter(cr_year == "2022", publisher == "Wiley", type == "ta_oa") |> pull(value) / publisher_summary_df |> filter(cr_year == "2022", publisher == "Wiley") |> pull(value) |> sum()*100`% (n = `r publisher_summary_df |> filter(cr_year == "2022", publisher == "Wiley", type == "ta_oa") |> pull(value)`) of Wiley's open access articles could be linked to transformative agreements in 2022. 
By contrast, Elsevier published fewer than half of its open access articles through transformative agreements (n = `r publisher_summary_df |> filter(cr_year == "2022", publisher == "Elsevier", type == "ta_oa") |> pull(value) `; `r publisher_summary_df |> filter(cr_year == "2022", publisher == "Elsevier", type == "ta_oa") |> pull(value) / publisher_summary_df |> filter(cr_year == "2022", publisher == "Elsevier") |> pull(value) |> sum()*100`%).

@@

Table \ref{tab:subject_summary_table} presents a high-level overview of hybrid open access by AJCS subject area using fractionalised counting to account for journals belonging to more than one category. 
Between 2018 and 2022, most hybrid journals with at least one open access article could be attributed to the Social Sciences category, which also includes Arts and Humanities.
However, these journals published the fewest number of articles, whereas Physical Sciences journals recorded the most articles, followed by Health Sciences and Life Sciences. 
In terms of open access,  Physical Sciences journals accounted for more than one-third of the articles published in the five-year period, followed by Health Sciences, Social Sciences and Life Sciences.

@@

Figure \ref{fig:subject_panel} presents the relative growth of hybrid open access by subject area between 2018 and 2022. In particular, Social Sciences including Arts and Humanties journals accounted for the strongest growth in the five-years period from `r subject_trend_plot$data |> filter(cr_year == "2018", top_level == "Social Sciences") |> pull(prop) |> sum() * 100`% (n = `r subject_trend_plot$data |> filter(cr_year == "2018", top_level == "Social Sciences") |> pull(value) |> sum()`) to `r subject_trend_plot$data |> filter(cr_year == "2022", top_level == "Social Sciences") |> pull(prop) |> sum() *100`% (n = `r subject_trend_plot$data |> filter(cr_year == "2022", top_level == "Social Sciences") |> pull(value) |> sum()`), followed by the Life Sciences from `r subject_trend_plot$data |> filter(cr_year == "2018", top_level == "Life Sciences") |> pull(prop) |> sum() * 100`% (n = `r subject_trend_plot$data |> filter(cr_year == "2018", top_level == "Life Sciences") |> pull(value) |> sum()`) to `r subject_trend_plot$data |> filter(cr_year == "2022", top_level == "Life Sciences") |> pull(prop) |> sum() *100`% (n = `r subject_trend_plot$data |> filter(cr_year == "2022", top_level == "Life Sciences") |> pull(value) |> sum()`) , Health Sciences from `r subject_trend_plot$data |> filter(cr_year == "2018", top_level == "Health Sciences") |> pull(prop) |> sum() * 100`% (n = `r subject_trend_plot$data |> filter(cr_year == "2018", top_level == "Health Sciences") |> pull(value) |> sum()`) to `r subject_trend_plot$data |> filter(cr_year == "2022", top_level == "Health Sciences") |> pull(prop) |> sum() *100`% (n = `r subject_trend_plot$data |> filter(cr_year == "2022", top_level == "Health Sciences") |> pull(value) |> sum()`) and Physical Sciences from `r subject_trend_plot$data |> filter(cr_year == "2018", top_level == "Physical Sciences") |> pull(prop) |> sum() * 100`% (n = `r subject_trend_plot$data |> filter(cr_year == "2018", top_level == "Physical Sciences") |> pull(value) |> sum()`) to `r subject_trend_plot$data |> filter(cr_year == "2022", top_level == "Physical Sciences") |> pull(prop) |> sum() *100`% (n = `r subject_trend_plot$data |> filter(cr_year == "2022", top_level == "Physical Sciences") |> pull(value) |> sum()`). 
<!-- In particular, journals in the subject areas recorded large high rate (see supplement)-->
Growth in Social Sciences category can be largely attributed to transformative agreements. 
In 2022, two-thirds of open access articles (`r subject_ta_trend_plot$data |> filter(cr_year == "2022", top_level == "Social Sciences") |> pull(prop) * 100`%, n = `r subject_ta_trend_plot$data |> filter(cr_year == "2022", top_level == "Social Sciences") |> pull(ta_oa_frac)`) were published by the first authors affiliated with participating institutions (see \ref{fig:subject_panel}B).

@@

Figure \ref{fig:subject_panel}C shows that this trend was consistent across hybrid journals belonging to the ASJC Social Sciences category.
In 2022, 25% of Social Sciences journals provided open access to at least every fourth article exclusively through transformative agreements.
However, hybrid open access through transformative agreements played a comparably lesser role in Life Sciences and Health Sciences. 
In these two subject areas, only about half of the open access articles can be linked to these agreements, both overall and on a median average across journals.
In contrast, the majority of Physical Science journals show an increase of open access through transformative agreements compared to other options to publish open access in hybrid journals.

```{r coutnry_data_prep}
country_all <- hoaddata::jn_aff |>
  filter(issn_l %in% hoaddata::jn_ind$issn_l) |>
  filter(!cr_year %in% c(2017, 2023), issn_l != "0027-8424") |>
  # only with cc
  filter(issn_l %in% hoaddata::cc_articles$issn_l) |>
  distinct(issn_l, cr_year, country_code, articles_total) |>
  group_by(cr_year, country_code) |>
  summarise(all = sum(articles_total))

country_oa <- hoaddata::cc_articles |>
  filter(issn_l %in% hoaddata::jn_ind$issn_l) |>
  filter(!cr_year %in% c(2017, 2023), issn_l != "0027-8424") |>
  group_by(cr_year, country_code) |>
  summarise(oa = n_distinct(doi))

country_codes <- readr::read_csv(here::here("data", "ror_country_codes.csv"))
country_ta <-  ta_oa_inst |>
  filter(issn_l %in% hoaddata::jn_ind$issn_l) |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  left_join(country_codes) |>
  group_by(cr_year, country_code) |>
  summarise(ta_oa = n_distinct(doi))

@@

```{r, country_plot, fig.asp=0.8}
country_plot <- ggplot(top_20_plot_df, aes(
    gsub("^20", "'", as.character(cr_year)),
    value / all_total,
    group = type,
    fill = type
  )) +
  geom_area(color = "#000000",
            linewidth = 0.15,
            stat = "identity") +
scale_fill_manual(
    NULL,
    values = c(no_ta = "#b3b3b3a0", ta_oa = "#56B4E9"),
    labels = c(no_ta = "Other", ta_oa = "Agreement"),
    guide = guide_legend(reverse = TRUE)
  ) +
  facet_wrap( ~ country_name_fct,
              nrow = 3,
              labeller = as_labeller(country_facet_labels)) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1L),
    expand = expansion(mult = c(0, 0.05)),
    breaks = scales::breaks_extended(4)
  ) +
  theme_minimal() +
  labs(y = "Hybrid OA Uptake", x = "2018 - 2022") +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.direction = "horizontal",
    axis.text.x = element_blank(),
    panel.grid.minor = element_blank(),
                panel.border = element_rect(color = "grey50", fill = NA),
    axis.title=element_text(size = 10),
    strip.text = ggtext::element_markdown(size = 6),
    plot.subtitle = ggtext::element_markdown(hjust = 0.5),
    panel.spacing = unit(1, "lines")
  )
```

@@

Between 2018 and 2022, high-income countries almost exclusively dominated hybrid open access publishing through transformative agreements. 
During this period, first authors affiliated with institutions from Organisation for Economic Co-operation and Development (OECD) member countries published `r oecd_stats_global |> filter(cat == "OECD") |> pull(oa)` open access articles in hybrid journals, representing `r oecd_stats_global |> filter(cat == "OECD") |> pull(oa) / n_oa * 100`% of the investigated open access articles.
This disparity between OECD nations and other countries becomes even more evident when considering open access through transformative agreements, as  `r oecd_stats_global |> filter(cat == "OECD") |> pull(ta_oa)` of `r n_ta_oa`, or `r oecd_stats_global |> filter(cat == "OECD") |> pull(ta_oa) / n_ta_oa*100`% of open access articles were associated with such agreements.
